Predicting the effect of candidate interventions for mitigating disease spread in asymmetric-mixing metapopulations

### 2024-07-22 ms.tex
wrote formalism paragraph, on to sketch of computational experiment

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.1in,letterpaper]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\title{Predicting the effect of candidate interventions for mitigating
disease spread in asymmetric mixing metapopulations}
\author{Matthew Adam Turner}
\date{}

@@

{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{Figures/ReserveMap.pdf}

\caption{Map of Dourados Indigenous Reserve villages (Bororó in light
blue, Jaguapiru in cyan) and the city of Dourados (dark blue). Note
Jaguapiru shares a much longer boundary with the reserve than Dourados,
which facilitates mobility in and out of the reserve for both Reserve
and city residents. Maps combined from Schnaufer et al.~(2023) and
Cavalcante (2019)}

\end{figure}%%
\begin{figure}[H]

{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{Figures/Time Series.pdf}

\caption{Time series of COVID-19 spread through the Dourados Indigenous
Reserve, with significant events highlighted. The first B.1.1 variant
cases were introduced by the Guarani ethnicity (1), followed about one
month later by a rapid rise in cases in Jaguapiru among the Terena
ethnicity(2); several months later a rapid rise in Bororó was caused by
a rapid rise among Guarani (3). Zeta was introduced at the end of 2020
(4), with a rapid rise around the holiday season especially in Jaguapiru
among Guarani (5); later in February cases rose in Bororó among the
Terena (6).}

@@

\section{Model}\label{model}

@@

{\centering \includegraphics{Figures/ModelSketch.pdf}

@@

\caption{We can model the relevant groups for mixing as either village
residence or ethnicity. Transmission, mixing, and mobility parameters
would be set accordingly. The \(C_g\) parameters would be reduced if
economic aid were allocated to keep group \(g\) on the Reserve during
outbreaks.}

@@

\subsection{Equations}\label{equations}

\begin{align*}
  \frac{dS_g}{dt} &= -\left(\tau c_{gg'} I_g + c_{g'g}I_{g'} + \gamma_g \right)S_g + r R_g   \\
  \frac{dI_g}{dt} &= \left(\tau c_{gg'} I_g + c_{g'g}I_{g'} + \gamma_g \right)S_g - \rho I_g \\
  \frac{dR_g}{dt} &= \rho I_g - r R_g
    \numberthis \label{eq:DiseaseContagion}
\end{align*}
\noindent

@@

{\centering \includegraphics[width=0.95\textwidth,height=\textheight]{Figures/ResultsSketch.pdf}

@@

\caption{Global reduction in travel to and from the city could have
differential effects on reduction in incidence for different groups.}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
% \@ifundefined{KOMAClassName}{% if non-KOMA class
  % \IfFileExists{parskip.sty}{%
  %   \usepackage{parskip}
  % }{% else
  %   % \setlength{\parindent}{pt}
  %   \setlength{\parskip}{6pt plus 2pt minus 1pt}}
% }{% if KOMA class
  % \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.1in,letterpaper]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\title{Predicting the effect of candidate interventions for mitigating
disease spread in asymmetric-mixing metapopulations}
\author{Matthew Adam Turner}
\date{\today}

@@

\mt{IF TERENA ARE MINORITY, HAVE HIGH MOBILITY TO CITY, AND $c_{GT} > c_{TG}$
THERE COULD BE A MINORITY INCUBATOR-TYPE EFFECT~\autocite{Turner2023}}

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:Map} ABOUT HERE ]} \\
\end{center}

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:DeOliveiraSeries} ABOUT HERE ]} \\
\end{center}

@@

\section{Model}\label{sec:model}

To understand and predict how different candidate economic interventions could
reduce transmission in the Dourados Reserve, we develop a mechanistic model of
transmission in the Reserve implemented as a set of compartmental
susceptible-infected-recovered system (SIR) of differential equations.  There
are six compartments in the differential equations, one $S_g$, $I_g$, and $R_g$ for
each group, $g$.  We evaluate different interventions by varying the reduction in
Indigenous travel between the Dourados reserve and the city, which in the real
world would be achieved by economic aid that reduces the need for Indigenous
travel to the city. First we test the effect of global reduction in travel on
infections in the Reserve, where aid is distributed equally across all possible
groupings. Second, we test different strategies for distributing aid to
different groups to understand which, if any, are more effective than equal aid
distribution. Finally, we test infectin prevention under both global and
group-specific intervention strategies for a stylized version of the Zeta
variant that allows probabilistically for re-infection. Although the model is
focused on COVID-19 transmission in the Dourados Indigenous Reserve, the
model is abstract. What we learn in our analyses could therefore transfer to
other diseases in the Reserve or other socio-epidemiological contexts.

\subsection{Model dynamics and formalism}
\label{sub:modelDynamics}

Our model represents interactions within and between Dourados Indigenous 
Reserve groups, and with members of the city of Dourados 
(Figure~\ref{fig:ModelSketch}). To represent
interactions between groups we define two possible groupings, by village
(Figure~\ref{fig:ModelSketch}, Model 1) and by ethnicity
(Figure~\ref{fig:ModelSketch}, Model 2). The within-group contact rate 
is $c_{gg}$ for focal group $g$. The contact rate between the focal group and
its outgroup, $g'$, is $c_{gg'}$. If in Model 2 the focal group is Terena, then
the within-group contact rate is $c_{TT}$ and the between-group contact rate,
representing how frequently Terena interact with Guarani is $c_{TG}$, Each of these groupings can be further
subdivided by the alternative Model grouping, represented in 
Figure~\ref{fig:ModelSketch} by the inset pie chart diagrams in each group's
circle. Reserve resident mobility is represented by a contact-infection rate,
$\gamma_g$, which represents how frequently a susceptible member of group $g$
brings an infection into the Reserve. The exact compartments, $S$, $I$, and $R$
for the city population are ignored.

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:ModelSketch} ABOUT HERE ]} \\
\end{center}

This contact structure, combined with a global transmission rate $\tau$,
recovery rate, $\rho$, and re-susceptibility rate, $r$, results in a system of
six differential equations for each of the six compartments
(Equation~\ref{eq:DiffEqs}). $\tau$ and $\rho$ will be set to reasonable values
for main analyses, but varied for sensitivity analyses.  \mt{Q: What constitutes
a ``reasonable value'' here? Values Aja used were for ABM---what about for a
diffeq model like this?}. To model increased re-infections observed with the
Zeta variant, we also include a term $r R_g$ that is positive in the
$\frac{dS_g}{dt}$ equation and negative in the $\frac{dR_g}{dt}$ term to
transfer recovered individuals from the recovered compartment for group $g$ to
the susceptible compartment for that group. If $r=0$, then there will be no
re-infections. Otherwise there may be re-infections, which could have complex
interactions with interaction structure.

\begin{align*}
  \frac{dS_g}{dt} &= -m_{gg'} \tau S_g + r R_g   \\
  \frac{dI_g}{dt} &=  m_{gg'} \tau S_g - \rho I_g \\
  \frac{dR_g}{dt} &= \rho I_g - r R_g,\quad\text{ where } \\[1em]
          m_{gg'} &= \left(c_{gg'} I_g + c_{g'g}I_{g'} + \gamma_g \right)
    \numberthis \label{eq:DiffEqs}
\end{align*}
\noindent

\vspace{0.5em}
\begin{center}
{[ TABLE \ref{tab:Variables} ABOUT HERE ]} \\
\end{center}

\subsection{Computational analysis}

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:GlobalTravelReduction} ABOUT HERE ]} \\
\end{center}

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:GroupSpecificTravelReduction} ABOUT HERE ]} \\
\end{center}

\begin{itemize}
  \item 
    Application to other problematic diseases in the Reserve, e.g., TB, STDs,\ldots 
  \item
    The data is not necessarily representative and we had to use educated guesses
    for mixing parameters.
  \item
    Have not yet considered more complex epidemiological and behavioral
    dynamics. For example, group structure can affect the coupled contagion of
    disease and behaviors that could encourage or impede the spread of
    disease~\autocite{SmaldinoJones2021}. Our compartmental model
    incorporated group structure and contact rate heterogeneity, but finer
    details of the social network structure can induce complex
    epidemiological dynamics~\autocite{Arthur2017}. In the case of Dourados
    and possibly other Indenous communities is the tendency for several
    generations of families to live in close proximinty to one another,
    often communally with other families as well, making social isolation
    culturally unpalatable and, consequentially, unfeasible. Adaptation to
    evolving, uncertain epidemiological social and environmental contexts is
    one further factor that can complicate epidemiological
    dynamics~\autocite{Arthur2021} and willingness to learn beneficial public
    health behaviors from others~\autocite{Turner2023a}.
\end{itemize}

\clearpage

\section{Tables}

\begin{table}[h]
  \centering
  \caption{\textbf{Model variables, description and values taken.} The
  compartmental model described in Equation~\ref{eq:DiffEqs} includes several
social mixing, behavioral, and epidemiological variables described here.}
  \label{tab:Variables}
  \begin{tabular}{rlp{1.3in}} 
    \toprule
    variable & description & values \\ 
    \midrule 
    $S_g,I_g,R_g$ & Susceptible, infected, recovered compartments by group & $[0.0,1.0]$ \\
    $c_{gg'},c_{g'g}$ & Asymmetric inter-group contact rates & $[0.0,1.0]$ \\
    $\gamma_g$ & Group-level incidence-contact rate with city residents &
      $\{0.0,0.1,\ldots,0.4\}$ \\
    $\rho$ & Global recovery rate & $[0.0,1.0]$ \\
    $\tau$ & Global ransmission rate & $[0.0,1.0]$ \\
    $r$    & Re-susceptible probability, non-zero only for Zeta & $0.0,0.5$ \\
    $g,g'$ & Reference group ($g$) and out-group $g'$ & Terena, Guarani; Bororó, Jaguapiru \\
    \bottomrule
  \end{tabular} 
\end{table}

\section{Figures}

@@

\centering
\caption{\textbf{Map of Dourados Indigenous Reserve villages.} Bororó is light
blue, Jaguapiru cyan, and the city of Dourados dark blue. Note
Jaguapiru shares a much longer boundary with the reserve than Dourados,
which facilitates mobility in and out of the reserve for both Reserve
and city residents. Maps combined from Cavalcante (2019) for the underlying
topographic map and Schnaufer et al.~(2023) for the villages, city, and 
sample collection points (clinics) layer.}
\label{fig:Map}
\vspace{0.3em}
{\includegraphics[width=0.8\textwidth,height=\textheight]{Figures/ReserveMap.pdf}}
\end{figure}%%

\begin{figure}[H]
\caption{Time series of COVID-19 spread through the Dourados Indigenous
Reserve, with significant events highlighted. The first B.1.1 variant
cases were introduced by the Guarani ethnicity (1), followed about one
month later by a rapid rise in cases in Jaguapiru among the Terena
ethnicity (2); several months later a rapid rise in Bororó was caused by
a rapid rise among Guarani (3). Zeta was introduced at the end of 2020
(4), with a rapid rise around the holiday season especially in Jaguapiru
among Guarani (5); later in February cases rose in Bororó among the
Terena (6).}
\label{fig:DeOliveiraSeries}
\vspace{0.3em}
{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{Figures/Time Series.pdf}

}
\end{figure}%

\caption{
  We can model the relevant groups for mixing as either village
residence or ethnicity. Mixing, and mobility parameters
would be set accordingly. Mixing is asymmetric, with $c_{gg}$ representing the
contact rate within group $g$, and each $c_{gg'}$ representing
the contact rate at which a member of group $g$ encounters a member of the
outgroup $g'$, e.g., $c_{GT}$ is the contact rate at which Guarani encounter
members of the Terena. The $\gamma_g$ parameters would be reduced through
economic aid to group $g$ for them to remain on the Reserve during
outbreaks. \mt{Looks like some labels are cut off; will fix in later draft.}}
\label{fig:ModelSketch}
\vspace{0.3em}
{\centering \includegraphics{Figures/ModelSketch.pdf}

}
\end{figure}%

\caption{Global reduction in travel to and from the city could have
Infection reduction ($\Delta I$, y-axis) for different groups could be
different as global reduction in travel to and from the city decreases
($\Delta \gamma$, x-axis).}
\label{fig:GlobalTravelReduction}
\vspace{0.3em}
{\centering \includegraphics[width=0.95\textwidth,height=\textheight]{Figures/ResultsSketch_ReductionByGroup.pdf}

@@

@@

@@

\caption{We hypothesize that overall reduction in disease incidence ($\Delta I$, y-axis)  depends on 
strategy to distribute the total aid to reduce travel between Dourados Reserve and city ($\Delta \gamma$, x-axis).}
\label{fig:GroupSpecificTravelReduction}
\vspace{0.3em}
{\centering \includegraphics[width=0.95\textwidth,height=\textheight]{Figures/ResultsSketch_ByStrategy.pdf}

@@
